jstor: Parsing XML - Lessons From Writing a Package

### 2018-10-01 draft.Rmd
expand draft with lessons learned and options to contribute

# Step 1: Hacking Away
Back in March 2017, I was starting out as a MA-student of
sociology in a research project
concerned with the scientific elite within sociology and economics. The project
had many goals, but writing an R package was not one of them. At the beginning
of my engagement,
I was presented with a dataset which was huge, at least for my terms: 
around 30GB of data, half of which was text, the other half 500,000 `.xml`-files.
The dataset was incredible in its depth: we basically sat on all articles
from JSTOR which belonged to the topics "sociology" and "economics". To repeat:
all articles that JSTOR has on those topics for all years.

@@

# Step two: doing it properly
Rewriting my functions was not that much of a hassle, in the end. I had turned
my functions into a package early on, and had already included many test cases
with [testthat](https://github.com/r-lib/testthat)
to make sure everything works as expected. This helped a lot for re-structuring
the code, since I already knew what my output should look like, and I only had 
to change the steps in between.

@@

# Lessons learned:

- community is important
- what could I have done differently?
Decision for API: each function is self-contained. If I had extracted the parts
for import somehow, it would be faster to apply different functions (like for
metada and references and other stuff).

# Acknowledgements

# Options to contribute

>>>>

# Hacking Away
Back in March 2017, I was starting out as a MA-student of
sociology in a research project
concerned with the scientific elite within sociology and economics. The project
had many goals, but writing an R package was not one of them. At the beginning
of my engagement,
I was presented with a dataset which was huge, at least for my terms: 
around 30GB of data, half of which was text, the other half 500,000 `.xml`-files.
The dataset was incredible in its depth: we basically sat on all articles
from JSTOR which belonged to the topics "sociology" and "economics". To repeat:
all articles that JSTOR has on those topics for all years.

@@

# Doing It Properly
Rewriting my functions was not that much of a hassle, in the end. I had turned
my functions into a package early on, and had already included many test cases
with [testthat](https://github.com/r-lib/testthat)
to make sure everything works as expected. This helped a lot for re-structuring
the code, since I already knew what my output should look like, and I only had 
to change the steps in between.

@@

# Lessons learned
I have learned many things while working on this package. While I aquired 
certain skills (like writing simple XPATH-queries), I want highlight a few 
general things.

Something that is true probably for many people working with and developing for
R, is that community is important. Without the efforts of many others, through
paid work or by spending their time voluntarily, developing the package would
not have been possible. This is true for the many packages my code builds on,
for the suite of packages that helps in developing and maintaining packages,
for answers I got over StackOverflow, and last but not least for the helpful
comments I received by @elinw and @jsonbecker during the review process for
onboarding the package.

A lesson which is more or less obvious from what I wrote above would be to 
benchmark your code, if you are planning on running it often. Packages like
`microbenchmark`, `profvis` or `bench` can help you in different ways to make 
sure that your code runs more efficiently.

At the beginning of the post I mentioned, that there would probably be a better
approach on parsing those files altogether. Before I finish, I want to briefly
elaborate on that thought.

The general approach I took when writing the package was inspired by the idea of
functional programming, that can be implemented in R through `lapply` or similar
functions within the `purrr`-package. The approach is to write a function that
solves your problem for one case, and then to apply it to all cases. In my case,
this leads to some duplication and inefficiency: The package has serveral 
functions that extract certain parts of the metadata-files. This makes sense,
since you only might be interested in certain parts, and parsing everything else
would be a waste of time. Unfortunately, if you happen to be interested in two
parts all the time, this means that each function has to read the original file
separately. There is no economy of scales, so to speak, because the file is
read again for each part which is to be extracted. An alternative would have 
been to write a general wrapper function that reads in the files and then 
applies each function in turn. I suspect that this would be quite complex, given
that the code should also be able to run in parallel in a proper way. This, and
the fact that I already spent *a lot* of time writing the package, mean that
I will most likely not add improvements on this side. There are however many 
other options for improvement, and I will gladly point you to a few of them.

# Options to contribute
I have strived for a proper test coverage, and I'd say it is decent at 91%.
There are a few cases where adding coverage would not be too difficult, which I
have mentioned in 
[issue #71](https://github.com/ropensci/jstor/issues/71). Another area
where there is still some work is in adding a few more fields which could be
extracted from the documents. This would involve some XPATH, but could be a good
starting point if you are curious about how those work. The corresponding issues
are [#23](https://github.com/ropensci/jstor/issues/23)
and [#32](https://github.com/ropensci/jstor/issues/32).
Any help, even fixing typos in the vignettes or documentation, is greatly 
appreciated, so if you want to get into contributing to a package, just go for 
it! I will help you with any hickups along the way, especially with questions
regarding Git, GitHub and pull requests.
